---
title: "STICS example in R"
author: "Timothée Flutre"
affiliation: "INRAE - GQE"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 5
    toc_float: true
    number_sections: TRUE
    code_folding: show
vignette: >
  %\VignetteIndexEntry{STICS example in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
params:
  eval_vignette: "load" # "run" means "(re-)run everything" / "load" means "load precomputed variables"
  path_to_JavaStics: "/home/tflutre/Documents/travail/src_ext/STICS/JavaSTICS-10.5.0-STICS-10.5.0"
  result_path: "/home/tflutre/Documents/travail/src/SticsRPacks/vignettes"
  precomputedFile: "STICS-example_precomputed.rds"
---

<!--
Steps to deploy this vignette:
1) set "eval_vignette" to "run" when editing the vignette so that any new precomputed variable
  is indeed saved into the rds file in inst/extdata
2) keep it like that for R CMD build and INSTALL
3) set "eval_vignette" to "load" and check that the vignette compiles properly
-->

```{r, eval=TRUE, include=FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = "#>"
)

precomputed <- list()
if (params$eval_vignette == "load") {
  file_path <- system.file(file.path("extdata", params$precomputedFile),
                           package = "SticsRPacks")
  stopifnot(file.exists(file_path))
  precomputed <- readRDS(file_path)
}

javastics <- params$path_to_JavaStics

## Useful function to work with params$eval_vignette:
#' @param var_names character vector of variable names
#' @param precomputed list of precomputed variables
#' @param eval_vignette use get() if "run", assign() if "load"
get_or_assign_var <- function(var_names, precomputed, eval_vignette, envir){
  stopifnot(eval_vignette %in% c("run", "load"))
  for(var_name in var_names){
    if (eval_vignette == "run"){
      precomputed[[var_name]] <- get(var_name, envir = envir)
    } else if (eval_vignette == "load"){
      assign(var_name, precomputed[[var_name]], envir = envir)
    }
  }
  return(precomputed)
}
```

```{r}
library(SticsRPacks)
library(ggplot2)
```


# Overview

As explained in the documentation available on the main [website](https://stics.inrae.fr/eng), the STICS acronym stands for "Scientific, Technical and Interdisciplinary simulator of soil-Crop System functioning".
More specifically, STICS is a generic soil-crop to support agro-ecosystem analysis, evaluation and design.
It runs at a daily time-step and uses input variables related to climate, soil and the cropping system and its management.
It can be described as a deterministic process-based model as well as a cropping system model.

The model equations are implemented in Fortran, and a graphical interface named JavaSTICS is coded in Java.
Both software are freely available for [download](https://stics.inrae.fr/eng/download) on the main website (after registration).
Moreover, to ease the programmatic use of STICS, a set of R packages, named [SticsRPacks](https://sticsrpacks.github.io/SticsRPacks/), has been developped.

The goal of this vignette is to **show how a new user, interested in STICS from an agronomic perspective, can quickly get familizarized with the model**, its required inputs and what outputs it can provide, **thanks to the functionalities from SticsRPacks**.
However, by no means should you think that following this vignette can dispense of reading the STICS book as well as JavaSTICS and SticsRPacks documentation... ;-)



# Set up the simulation

Concretely, this vignette will take a single example among all of those provided in the JavaSTICS "example" directory.
Below the "wheat" example is chosen, but feel free to copy-paste the code to another script and choose another example.

## Choose a USM

The USM acronym stands for "unit of simulation".

Let us list all the example USMs provided with JavaSTICS:

```{r, eval=params$eval_vignette == "run"}
allUsmsFile <- file.path(javastics, "example", "usms.xml")
usm_names <- get_usms_list(allUsmsFile)
```

```{r, echo=FALSE}
precomputed <- get_or_assign_var("usm_names", precomputed, params$eval_vignette, environment())
```

```{r}
length(usm_names)
usm_names
usm_name <- "wheat" # as a user, you may wish to choose another one, e.g., pea
stopifnot(usm_name %in% usm_names)
```

## Prepare the working directory

It is a good habit to start working from a new, clean directory, called a `workspace` in SticsRPacks.

```{r, eval=params$eval_vignette == "run"}
workspace <- file.path(tempdir(), paste0("vignette-STICS-example_", usm_name))
if (dir.exists(workspace)) {
  unlink(workspace, recursive = TRUE)
}
```

The inputs required for STICS are often available in the XML format.
All necessary XML files are listed and described in section 2.3 of the JavaSTICS documentation.

Here they will be retrieved from the JavaStics example directory:
```{r, eval=params$eval_vignette == "run"}
extract_workspace(from_workspace = file.path(javastics, "example"),
                  to_workspace = workspace,
                  javastics = params$path_to_JavaStics,
                  usm = usm_name)
usm_details <-
  get_param_xml(file.path(javastics, "example", "usms.xml"),
                select="usm", select_value=usm_name)[[1]]
```

```{r, echo=FALSE}
precomputed <- get_or_assign_var("usm_details", precomputed, params$eval_vignette, environment())
```

```{r}
usm_details
```

## Explore the input files

### Years

```{r, eval=params$eval_auto_vignette}
## TODO: include this function in SticsRFiles
get_usm_years <- function(workspace, usmSelVal, usmsFile="usms.xml"){
  stopifnot(dir.exists(workspace),
            length(usmSelVal) == 1,
            length(usmsFile) == 1,
            file.exists(file.path(workspace, usmsFile)))

  out <- c("year1"=NA, "year2"=NA)

  fclim1 <- get_param_xml(file=file.path(workspace, usmsFile),
                          param="fclim1", select="usm",
                          select_value=usmSelVal)[[1]]$fclim1
  line1 <- readLines(file.path(workspace, fclim1), n=1)
  tmp <- strsplit(line1, "\\s+")[[1]]
  out["year1"] <- as.numeric(tmp[2])
  out["year2"] <- out["year1"] + 1

  return(out)
}
```

```{r, eval=params$eval_vignette == "run"}
years <- get_usm_years(workspace, usm_name)
# if(packageVersion("SticsRFiles") >= "1.7.0")
#   years <- get_years(workspace, "usms.xml", usm_name) # as.vector(years[usm_name,])
```

```{r, echo=FALSE}
precomputed <- get_or_assign_var("years", precomputed, params$eval_vignette, environment())
```

```{r}
years
```

### Management

```{r, eval=params$eval_auto_vignette}
## TODO
```

### Phenology

An important phenological stage for small-grain cereals is called "IAMF" in STICS, that corresponds to the "one-centimeter ear" stage (BBCH 30).

In STICS, it can be a variable, in which case its value will be simulated:
```{r, eval=FALSE}
get_var_info("iamf")
```
```{r, echo=FALSE}
knitr::kable(get_var_info("iamf"))
```

But it can also be a parameter, in which case its value needs to be given before the simulation:
```{r, eval=FALSE}
get_param_info("iamf")
```
```{r, echo=FALSE}
knitr::kable(get_param_info("iamf"))
```

As you can see above in the "file" column, this parameter value can be found in the management file ("tec"):
```{r, eval=params$eval_vignette == "run"}
form_pheno_stages <- get_param_xml(file.path(workspace, usm_details$ftec[1]),
                                   select="formalisme",
                                   select_value="phenological stages")
```

```{r, echo=FALSE}
precomputed <- get_or_assign_var("form_pheno_stages", precomputed, params$eval_vignette, environment())
```

```{r}
unlist(form_pheno_stages)
```

The value can be directly retrieved (as a Julian day) and further converted into a calendar date:
```{r, eval=params$eval_vignette == "run"}
iamf <- unlist(get_param_xml(file.path(workspace, usm_details$ftec[1]), "iamf"))
```

```{r, echo=FALSE}
precomputed <- get_or_assign_var("iamf", precomputed, params$eval_vignette, environment())
```

```{r}
iamf
compute_date_from_day(day=iamf, start_year=years["year1"])
```

## Convert XML into TXT

The last step before running the simulation consists in converting the input XML files used by JavaSTICS into text files that STICS can use.

Currently, the workspace contains several files as well as a single directory, named `"plant"`, that with the species and genotype-specific parameters:
```{r, eval=params$eval_vignette == "run"}
list_wksp_0 <- list.files(workspace, recursive = TRUE)
```

```{r, echo=FALSE}
precomputed <- get_or_assign_var("list_wksp_0", precomputed, params$eval_vignette, environment())
```

```{r}
list_wksp_0
```

The following function will convert the XML files into their text equivalent:
```{r, eval=params$eval_vignette == "run"}
gen_usms_xml2txt(javastics, workspace, usm = usm_name)
```

As you can see below, the workspace now contains a new directory, named by the USM, which only contains text files:
```{r, eval=params$eval_vignette == "run"}
list_wksp_1 <- list.files(workspace, recursive = TRUE)
```

```{r, echo=FALSE}
precomputed <- get_or_assign_var("list_wksp_1", precomputed, params$eval_vignette, environment())
```

```{r}
list_wksp_1
```

## Choose output variables

In the conversion process, a file (compulsorily named `"var.mod"`) was also generated.
It contains a default list of variables that will be saved by STICS during the simulation.
```{r, eval=params$eval_vignette == "run"}
var_mod_0 <- readLines(file.path(workspace, usm_name, "var.mod"))
```

```{r, echo=FALSE}
precomputed <- get_or_assign_var("var_mod_0", precomputed, params$eval_vignette, environment())
```

```{r}
var_mod_0
```

If you are interested in other variables not included in this default list, you need to append them to this file.
For instance, if you are used to conduct field trials, you may want variables that are commonly phenotyped, such as several important developmental stages, as well as LAI/GAI, height, above-ground biomass and grain weight.
It is also the occasion to add more variables you are interested in and that STICS will simulate (variables already in the default list will not be added twice).

```{r}
devPars <- c("iplt", "iger", "ilev", "iamf", "ilax",
             "iflo","idrp","ilan","imat","irec")
dev_vars <- paste0(devPars, "s")
varsToAdd <- c(dev_vars,
               "lai_n", "leai", "hauteur", "masec_n", "mafruit",
               "mafeuil", "mafeuilverte",
               "matigestruc")
```

```{r, eval=params$eval_vignette == "run"}
gen_varmod(workspace,
           var = varsToAdd,
           file_name = file.path(usm_name, "var.mod"),
           append = TRUE)
```



# Run the simulation

```{r, eval=params$eval_vignette == "run"}
wrapper_options <- stics_wrapper_options(javastics, workspace = workspace, parallel = FALSE)
wrap <- stics_wrapper(wrapper_options)
```



# Explore the outputs

## Retrieve observed and simulated data

The simulation outputs will be saved as a list of data frame, one per USM (here, there is a single one).
The observations, if any, can be retrieved in a similar way.
```{r, eval=params$eval_vignette == "run"}
sim <- get_sim(workspace, usm = usm_name)
obs <- get_obs(workspace, usm = usm_name)
```

```{r, echo=FALSE}
precomputed <- get_or_assign_var(c("sim","obs"), precomputed, params$eval_vignette, environment())
```

```{r}
str(sim, list.len = 10)
str(obs)
```

## Plot observed and simulated data

```{r}
for(out_var in c("masec_n", "hauteur", "lai_n")){
  p <- plot(sim, var = out_var, obs = obs)[[usm_name]] +
    theme_bw()
  print(p)
}
```

## Add developmental stages

First we need to retrieve the dates of the developmental stages we are interested in:
```{r, eval=params$eval_auto_vignette}
## TODO: see issue to include this function in SticsRFiles
get_dev_stages <- function(sim,
                           dev_vars=c("iplts", "igers", "ilevs", "iamfs",
                                      "ilaxs", "iflos", "idrps", "ilans",
                                      "imats", "irecs"),
                           usm=NULL){
  #get_var_info(keyword="date")

  stopifnot(is.list(sim),
            ! is.null(names(sim)))

  out <- list()

  if(is.null(usm))
    usm <- names(sim)
  stopifnot(all(usm %in% names(sim)))

  for(usm_name in usm){
    dat <- sim[[usm_name]]
    firstYear <- as.integer(sort(unique(format(dat$Date, format="%Y")))[1])
    out_usm <- data.frame(var=dev_vars,
                          day=NA,
                          date="")
    for(i in 1:nrow(out_usm)){
      dev_var <- dev_vars[i]
      if(! dev_var %in% colnames(dat)){
        msg <- paste0("'", dev_var, "' not found")
        warning(msg, immediate. = TRUE)
        next
      }
      tmp <- table(dat[[dev_var]])
      if(length(tmp) == 1){ # ex. iplts
        out_usm$day[i] <- as.integer(names(tmp)[1])
      } else if(length(tmp) == 2){
        stopifnot(names(tmp)[1] == "0")
        out_usm$day[i] <- as.integer(names(tmp)[2])
      } else
        stop(paste0(dev_var, " has more than three values"))
      out_usm$date[i] <- as.character(compute_date_from_day(out_usm$day[i], firstYear))
    }
    out_usm$date <- as.Date(out_usm$date)
    out_usm$Date <- as.POSIXct(out_usm$date) # required to plot with CroPlotR
    out_usm$var <- factor(out_usm$var,
                          levels=out_usm$var[order(out_usm$day)],
                          ordered=TRUE)
    out[[usm_name]] <- out_usm
  }

  return(out)
}

(dev_stages <- get_dev_stages(sim, dev_vars))
# if(packageVersion("SticsRFiles") >= "1.7.0")
#   dev_stages <- get_dev_stages(workspace, "usms.xml", usm_name)[[usm_name]]
```

And then we can add them to the plot:
```{r, eval=params$eval_auto_vignette}
for(out_var in c("masec_n", "hauteur", "lai_n")){
  p <- plot(sim, var = out_var, obs = obs)[[usm_name]] +
    geom_vline(data=dev_stages[[usm_name]], aes(xintercept=Date, color=var)) +
    theme_bw()
  print(p)
}
```



# Play with the inputs

## Change the sowing density

### Run the simulations

```{r}
get_param_info(keyword="sowing density")
```

As indicated in the "file" column, the "tec" file needs to be modified.
We will explore the range of possible sowing densities, and run STICS for each of them.

```{r}
new_densities <- unique(c(seq(5, 50, by=5),
                          seq(50, 1000, by=25))) #, seq(1000, 2000, by=100)))
param_values <- data.frame(densitesem = new_densities)#[1:10,,drop=F]))
```

```{r, eval=params$eval_vignette == "run"}
wrapper_options <- stics_wrapper_options(javastics, workspace = workspace)
system.time(
    wrap_densitesem <- stics_wrapper(wrapper_options, param_values))
## TODO: la commande ci-dessous va evoluer, issue en cours
wrap_densitesem$sim_list[[usm_name]]$densitesem <- 
  rep(param_values$densitesem,
      each = length(unique(wrap_densitesem$sim_list[[usm_name]]$Date)))
```

```{r, echo=FALSE}
precomputed <- get_or_assign_var("wrap_densitesem", precomputed, params$eval_vignette, environment())
```

### Extract the biomass at harvest

```{r}
colsToKeep <-  c("densitesem", "Date", "irecs", "masec_n", "mafruit")
datCFY <- wrap_densitesem$sim_list[[usm_name]][, colsToKeep]
rowsToKeep <- c()
for(densitesem in unique(sort(datCFY$densitesem))){
  is_densitesem <- (datCFY$densitesem == densitesem)
  harvest_Julian <- as.integer(names(table(datCFY[is_densitesem, "irecs"]))[2])
  rowsToKeep <- c(rowsToKeep,
                  which(is_densitesem & datCFY$irecs == harvest_Julian)[1])
}
datCFY <- datCFY[rowsToKeep,]
rownames(datCFY) <- NULL
head(datCFY)
```

### Plot the "law" of constant final yield

```{r, fig.width=12}
ggplot(datCFY) +
  aes(x=densitesem, y=masec_n) +
  geom_smooth(formula=y~x, method="loess", span=0.4) +
  geom_point(size=2) +
  scale_x_continuous(breaks=seq(0, 2000, by=100)) +
  labs(title=usm_name) +
  theme_bw()
```



```{r, eval=params$eval_vignette == "run", echo=FALSE}
output_dir_path <- file.path(getwd(), "..", "inst", "extdata")
stopifnot(dir.exists(output_dir_path))
output_file_path <- file.path(output_dir_path, params$precomputedFile)
if (file.exists(output_file_path))
  tmp <- file.remove(output_file_path)
saveRDS(precomputed, output_file_path)
```

```{r, echo=FALSE}
knitr::knit_exit()
```



## Change the nitrogen fertilization

```{r}
## TODO
```
