---
title: "STICS example in R"
author: "Timothée Flutre"
affiliation: "INRAE - GQE"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 5
    toc_float: true
    number_sections: TRUE
    code_folding: show
vignette: >
  %\VignetteIndexEntry{STICS example in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
params:
  eval_auto_vignette: TRUE     # automatic execution of the vignette
  eval_manual_vignette: FALSE  # manual execution of the vignette (e.g. to 
                               # create results graphs included in it)
  eval_auto_test: FALSE  # automatic test execution of the vignette, SET IT TO
                         #  TRUE IF manual_vignette IS SET TO TRUE
  path_to_JavaStics: "/home/tflutre/Documents/travail/src_ext/STICS/JavaSTICS-10.4.1-STICS-10.4.1"
  result_path: "/home/tflutre/Documents/travail/src/SticsRPacks/vignettes"
---

```{r, eval=TRUE, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
javastics <- params$path_to_JavaStics
```

```{r}
library(SticsRPacks)
library(XML) # TODO: see below
## TODO: save necessary variables as a list in a file inst/extdata/...rds
##   to display the results on GitHub that cannot re-run the vignette
```


# Overview

As explained in the documentation available on the main [website](https://stics.inrae.fr/eng), the STICS acronym stands for "Scientific, Technical and Interdisciplinary simulator of soil-Crop System functioning".
More specifically, STICS is a generic soil-crop to support agro-ecosystem analysis, evaluation and design.
It runs at a daily time-step and using input variables related to climate, soil and the cropping system and its management.
It can be described as a deterministic process-based model as well as a cropping system model.

The model equations are implemented in Fortran, and a graphical interface named JavaSTICS is coded in Java.
Both software are freely available for [download](https://stics.inrae.fr/eng/download) on the main website.
Moreover, to ease the programmatic use of STICS, a set of R packages, named [SticsRPacks](https://sticsrpacks.github.io/SticsRPacks/), has been developped.

The goal of this vignette is to **show how a new user, interested in STICS from an agronomic perspective, can quickly get familizarized with the model**, its required inputs and what outputs it can provide, **thanks to the functionalities from SticsRPacks**.
However, by no means should you think that following this vignette can dispense of reading the STICS book as well as JavaSTICS and SticsRPacks documentation... ;-)



# Make a workspace

Concretely, this vignette will take an example among all of those provided in the JavaSTICS "example" directory.
Below the "wheat" example is chosen, but feel free to copy-paste the code to another script and choose another example.

## Choose a USM

The USM acronym stands for "unit of simulation":
```{r, eval=params$eval_auto_vignette}
allUsmsFile <- file.path(javastics, "example", "usms.xml")
usmNames <- get_usms_list(allUsmsFile)
length(usmNames)
usmName <- "wheat" # as a user, you may wish to choose another one
stopifnot(usmName %in% usmNames)
```

## Create a working directory

It is a good habit to start working from a new, clean directory, called a `workspace` in SticsRPacks.

```{r, eval=params$eval_auto_vignette}
workspace <- file.path(tempdir(), paste0("vignette-STICS-example_", usmName))
if(dir.exists(workspace))
  unlink(workspace, recursive=TRUE)
dir.create(workspace)
```



# Retrieve all necessary files

They are listed and described in section 2.3 of the JavaSTICS documentation.

## Make `usms.xml`

To run a simulation, it is compulsory to have such a file, with this specific name, in the workspace.
This XML file contains some parameter values as well as the path to input files required for the simulation.

```{r, eval=params$eval_auto_vignette}
## TODO: use a function from SticsRFiles to avoid using the XML package directly
doc <- xmlParse(allUsmsFile)
root <- xmlRoot(doc)
listUSMs <- xmlChildren(root)
length(listUSMs)
usmNames <- xmlSApply(root, xmlGetAttr, "nom")
toKeep <- which(usmNames == usmName)
for(i in 1:length(listUSMs)){
  if(! i %in% toKeep)
    removeChildren(xmlParent(listUSMs[[i]]), listUSMs[[i]])
}
stopifnot(length(xmlChildren(root)) == 1)
saveXML(doc, file.path(workspace, "usms.xml"))
```

## Copy other input files

For STICS to run successfully, it requires several input files, whose list can be obtained with the following commands:
```{r, eval=params$eval_auto_vignette}
usm_details <-
  get_param_xml(file.path(workspace, "usms.xml"),
                select="usm", select_value=usmName)[[1]]
names(usm_details)
tmp <- get_param_info(names(usm_details))
tmp <- tmp[grep("^name of", tmp$definition),]; rownames(tmp) <- NULL
tmp[,c("name","definition")]
```

They can then be copied to the workspace:
```{r, eval=params$eval_auto_vignette}
file.copy(file.path(javastics, "example", usm_details$finit), workspace)
file.copy(file.path(javastics, "example", usm_details$fstation), workspace)
file.copy(file.path(javastics, "example", usm_details$fclim1), workspace)
file.copy(file.path(javastics, "example", usm_details$fclim2), workspace)
## the plant file should be in a subdirectory named "plant"
if(! dir.exists(file.path(workspace, "plant")))
  dir.create(file.path(workspace, "plant"))
file.copy(file.path(javastics, "plant", usm_details$fplt[1]),
          file.path(workspace, "plant"))
file.copy(file.path(javastics, "example", usm_details$ftec[1]), workspace)
file.copy(file.path(javastics, "example", usm_details$flai[1]), workspace)
file.copy(file.path(javastics, "example", usm_details$fobs[1]), workspace)
file.copy(file.path(javastics, "example", "sols.xml"), workspace)

list.files(workspace)
```

## Make `var.mod`

To run a simulation, it is compulsory to have such a file, with this specific name, in the workspace.
This text file contains the simulated variables to keep in the output.

```{r, eval=params$eval_auto_vignette}
devPars <- c("iplt", "iger", "ilev", "iamf", "ilax",
             "iflo","idrp","ilan","imat","irec")
devVars <- paste0(devPars, "s")

growthVars <- c("masec_n", "msrac_n", "masecveg", "mafruit",
                "mafeuil", "mafeuilverte",
                "matigestruc")

otherVars <- c("lai_n", "leai", "hauteur", "nbinflo_recal")

outVars <- c(devVars, growthVars, otherVars)

get_var_info(outVars)

gen_varmod(workspace, var=outVars)
```


# Run the simulation

```{r, eval=params$eval_auto_vignette}
gen_usms_xml2txt(javastics, workspace, usm=usmName)
wrapper_options <- stics_wrapper_options(javastics, workspace=workspace, parallel=FALSE)
wrap <- stics_wrapper(wrapper_options)
```



# Explore the outputs

## Retrieve observed and simulated data

```{r, eval=params$eval_auto_vignette}
sim <- get_sim(workspace, usm=usmName)
colnames(sim[[1]])

## (devStages <- getDevStages(sim, devVars))

obs <- get_obs(workspace, usm=usmName)
colnames(obs[[1]])
```

## Plot observed and simulated data

```{r, eval=params$eval_auto_vignette}
plot(sim, var="lai_n", obs=obs)
plot(sim, var="hauteur", obs=obs)
plot(sim, var="masec_n", obs=obs)
```
